PyPy's Approach to Virtual Machine Construction

### 2006-07-12 talk/dls2006/paper.tex
typo
adjustements in the intro

\begin{abstract}
The PyPy project seeks to prove both on a research and a
practical level the feasibility of writing a virtual machine (VM)
for a dynamic language in a dynamic language -- in this case, Python.
The aim is to translate (i.e.\ compile) the VM to arbitrary target
environments, ranging in level from C/Posix to Smalltalk/Squeak via
Java and CLI/.NET, while still being of reasonable efficiency within
these environments.

A key tool to achieve this goal is the systematic reuse of the
(unmodified, dynamically typed) Python language as a system
programming language at various levels of our architecture and
translation process.  For each level, we design a corresponding type
system and apply a generic type inference engine -- for example, the
garbage collector is written in a style that manipulates
simulated pointer and address objects, and when translated to C
these operations become C-level pointer and address instructions.
\end{abstract}

@@

However, we believe that platform standardization does not have to be
a necessary component of this equation.  We are basically using the
standard ``meta-programming'' argument: if one could write the VM in a
very high level language, then the VM itself could be automatically
\textit{translated} to any lower-level platform.  Moreover by writing
the VM in such a language we would gain in flexibility in
architectural choices and expressiveness.

PyPy achieves this goal without giving up on the efficiency of the
compiled VMs.

The key factors enabling this result are not to be found in recent
advances in any particular research area -- we are not for example using
constraint-based type inference.  Instead, we are following a novel
overall architecture: it is split into many levels of stepwise
translation from the high-level source of the VM to the final target
platform.  Similar platforms can reuse many of these steps, while for
very different platforms we have the option to perform very different
translation steps.  Each step reuses a common type inference component
with a different, ad-hoc type system.

@@

It is also simple, in the context of abstract interpretation, to
perform translation-time constant propagation, computations and
implement conditional code in a natural way. For example, the code in
figure \ref{fig_insertclean}, which is then specialized for the
possible types of dictionary \texttt{d}, shows this kind of
compile-time computation and introspection on low-level types,
\texttt{ENTRY} is a C-like structure type and the
\texttt{hasattr}-ibute checks introspect it at translation time. This
kind of expressiveness increased our ability of reuse code even in
low-level helpers.

>>>>

\begin{abstract}
The PyPy project seeks to prove both on a research and a practical
level the feasibility of constructing a virtual machine (VM) for a
dynamic language in a dynamic language -- in this case, Python.  The
aim is to translate (i.e.\ compile) the VM to arbitrary target
environments, ranging in level from C/Posix to Smalltalk/Squeak via
Java and CLI/.NET, while still being of reasonable efficiency within
these environments.

A key tool to achieve this goal is the systematic reuse of the
Python language as a system
programming language at various levels of our architecture and
translation process.  For each level, we design a corresponding type
system and apply a generic type inference engine -- for example, the
garbage collector is written in a style that manipulates
simulated pointer and address objects, and when translated to C
these operations become C-level pointer and address instructions.
\end{abstract}

@@

However, we believe that platform standardization does not have to be
a necessary component of this equation.  We are basically using the
standard ``meta-programming'' argument: if one could write the VM at a
sufficiently high level of abstraction in a language supporting such
level, then the VM itself could be automatically \textit{translated}
to any lower-level platform.  Moreover by writing the VM in such a way
we would gain in flexibility in architectural choices and
expressiveness.

PyPy achieves this goal without giving up on the efficiency of the
compiled VMs, or keeping it within reach of further reasonable
optimization work.

The key factors enabling this result are not to be found in recent
advances in any particular research area -- we are not for example
using constraint-based type inference.  Instead, we are following a
novel overall architecture: it is split into many levels of stepwise
translation from the high-level source of the VM to the final target
platform, each step adding support for features that were assumed
primitive by the previous ones.  Similar platforms can reuse many of
these steps, while for very different platforms we have the option to
perform very different translation steps.  Each step reuses a common
type inference component with a different, ad-hoc type system.

@@

It is also simple, in the context of abstract interpretation, to
perform translation-time constant propagation, computations and
implement conditional code in a natural way. For example, the code in
figure \ref{fig_insertclean}, which is then specialized for the
possible types of dictionary \texttt{d}, shows this kind of
compile-time computation and introspection on low-level types,
\texttt{ENTRY} is a C-like structure type and the
\texttt{hasattr}-ibute checks introspect it at translation time. This
kind of expressiveness increased our ability to reuse code even in
low-level helpers.
